\chapter{Discussion}
\label{chap:discussion}

This chapter discusses the key findings, design tradeoffs, and limitations of both proposed frameworks in light of the experimental results presented in Chapter~\ref{chap:experiments}.


%% ============================================================
\section{Discussion on BackFlush}\label{sec:disc_backflush}
%% ============================================================

\subsection{BackFlush-GA against BackFlush-RoPE Tradeoff}

The experimental results in Table~\ref{tab:defense_comparison} reveal a clear tradeoff between the two BackFlush variants. BackFlush-GA achieves marginally lower ASR on some configurations, for example 1.36\% against 0.98\% on Mistral-7B, and higher CACC at 99.68\% against 98.58\%, but consistently fails to preserve watermarks across all three architectures. BackFlush-RoPE achieves comparable or better ASR, below 1\% across all architectures, while preserving watermarks, at the cost of slightly lower CACC in some cases. The utility gap is decisive, as BackFlush-RoPE maintains perplexity close to the clean baseline at approximately 6, while BackFlush-GA degrades utility to 10--11 perplexity. This tradeoff confirms that the unbounded gradient signal in GA, while effective for backdoor disruption, causes indiscriminate parameter degradation that damages both watermark subspaces and general model utility. RoPE's bounded rotation loss $\mathcal{L}_{\textit{rotate}} \in [0, 2]$ provides sufficient signal to flush backdoors while constraining the update magnitude to preserve other parameter structures.

\subsection{Limitations of BackFlush}

\noindent
\textbf{Dependence on a clean baseline model.} The detection mechanism via Backdoor Susceptibility Amplification requires access to a clean baseline $\mathcal{M}_{\textit{clean}}$ for loss comparison. In practice, obtaining a verified clean model of the same architecture is feasible for popular open-source families such as Llama, Mistral, and Qwen, but may not generalize to proprietary or custom architectures where no trusted clean reference exists.

\medskip
\noindent
\textbf{Auxiliary data selection.} BackFlush uses SciQ as $\mathcal{D}_{\textit{aux}}$ in all experiments. While the Backdoor Flushing Phenomenon is demonstrated to be robust across trigger types (Table~\ref{tab:poison_config}), the sensitivity of the framework to the choice of auxiliary dataset has not been exhaustively characterized. Different auxiliary domains may entangle with backdoor subspaces to varying degrees, potentially affecting removal efficacy.

\medskip
\noindent
\textbf{SFT-only defense evaluation.} Although the threat model in Chapter~\ref{chap:threat_model} formalizes SFT-based, RLHF-based, and editing-based backdoor vectors, the experimental evaluation focuses on SFT-based poisoning. The generalization of the Backdoor Flushing Phenomenon to RLHF-corrupted reward models and editing-based weight transplants remains to be validated empirically.

\medskip
\noindent
\textbf{Detection threshold sensitivity.} The detection function $\mathcal{F}$ relies on a threshold $\tau$ to distinguish backdoored from clean models. While Figure~\ref{fig:backdoor_detection}(b) demonstrates consistent detection gaps across varying backdoor counts, the optimal setting of $\tau$ and its robustness to edge cases, such as models with very few or very subtle backdoors, requires further investigation.


%% ============================================================
\section{Discussion on RePAIR}\label{sec:disc_repair}
%% ============================================================

\subsection{Effectiveness of Training-Free Unlearning}

The single-sample ablation (Table~\ref{tab:single_sample}) provides the strongest validation of STAMP's design. All six training-based baselines fail completely when $|\mathcal{D}_f| = 1$, with $Acc_f$ remaining at 100, while STAMP and STAMP-LR achieve $Acc_f = 0.00$ with $Acc_r > 70$. This confirms that the gradient signal bottleneck is not a tuning problem but a fundamental limitation of gradient-based unlearning under single-sample conditions. STAMP's closed-form pseudoinverse update bypasses this bottleneck entirely, making it the only viable mechanism for the interactive setting where user requests arrive individually.

\subsection{Limitations of RePAIR}

\noindent
\textbf{Scope of the single-sample claim.} \textit{Single-sample} refers to the size of the forget set, $|\mathcal{D}_f| = 1$. Existing LLM unlearning methods need a batch of forget examples and fail at this granularity (Table~\ref{tab:single_sample}), whereas STAMP removes a target from one. This does not mean that no other data is used, since the retain buffer and the shared reference set $\mathcal{D}_{\textit{ref}}$ of 200 refusal prompts, fixed across all requests, are still required.

\medskip
\noindent
\textbf{Retain data at inference time.} Although STAMP operates with as little as a 10\% retain ratio (Table~\ref{tab:retain_ratio}), it still requires a small replay buffer $\mathcal{D}_r$. This dependence is not specific to STAMP, as every baseline compared against also relies on retain data, and retain-free unlearning is an emerging direction rather than a solved one. FLAT~\citeauthor{wang2024llm} is the notable exception, operating on $\mathcal{D}_f$ alone through a regularizer. Storing the buffer on an edge device sits uneasily with the GDPR and CCPA motivation on which the framework rests, since the user must keep a slice of the data the system is meant to remove.

\medskip
\noindent
\textbf{Three models rather than one.} RePAIR assigns intent detection, repair-code generation, and the edit itself to three separate models. This reflects the open-weight models available rather than a requirement of the method, since no single open model tested was at once a reliable intent classifier and a competent code generator. A model strong in both general reasoning and code would subsume all three roles, allowing $\mathcal{M}_{\textit{watchdog}}$ and $\mathcal{M}_{\textit{surgeon}}$ to be dropped and leaving a single model that repairs itself.

\medskip
\noindent
\textbf{Resource constraints at test time.} As shown in Table~\ref{tab:cost_comparison}, training-based methods require significantly more GPU memory than is typically available at inference. The STAMP-LR update itself is light, needing under 4 MB of workspace and no gradients, so the repair runs on device. The remaining cost is the surrounding pipeline, which currently keeps $\mathcal{M}_{\textit{watchdog}}$ and $\mathcal{M}_{\textit{surgeon}}$ resident alongside $\mathcal{M}_{\textit{patient}}$. Collapsing these into a single capable model, or serving them remotely, would leave only $\mathcal{M}_{\textit{patient}}$ in memory and bring the full framework within an on-device budget.

\medskip
\noindent
\textbf{Turnaround time.} End-to-end latency is 9.36 minutes for STAMP and 6.50 for STAMP-LR (Table~\ref{tab:repair_pipeline}), which is not what is usually meant by \textit{interactive}. The breakdown is informative: of the 9.36 minutes, 7.13 are the repair itself and 2.23 are orchestration, namely the calls to $\mathcal{M}_{\textit{watchdog}}$ and $\mathcal{M}_{\textit{surgeon}}$ together with model loading and I/O. For STAMP-LR the split is 4.25 and 2.25, so the orchestration overhead is essentially constant and independent of the unlearning method. Within the repair, the closed-form solve itself takes seconds, and almost all of the remaining time is activation collection, performed one sample at a time in the current implementation. Batching these forward passes and caching $\mathcal{D}_{\textit{ref}}$, which is fixed across requests, would remove most of this cost. The present latency is therefore a property of the implementation rather than of the method, and bringing it down to conversational speed is left to future work.

\medskip
\noindent
\textbf{Layer selection generality.} Table~\ref{tab:layer_ablation} identifies layer~6 of Llama-3-8B as the intervention point, following ROME and MEMIT's localization of factual recall to the early feed-forward layers~\citeauthor{meng2022locating,meng2022mass}. This selection is architecture-specific, as different models with different layer counts and representation structures may require re-running the layer sweep to identify the appropriate intervention layer.

\medskip
\noindent
\textbf{Abuse of the interface.} A natural-language unlearning channel is itself an attack surface, since the same request that removes a private fact could be used to remove refusal behaviour or safety-critical knowledge. This is not evaluated here, and access control and deployment guidelines are left to future work. The exposure is bounded by the setting targeted, namely a personal model whose knowledge resides in the weights, such as a standalone model provisioned for a single user who cannot write unlearning code. Hosted assistants instead hold personal details in the context window and answer from it, in which case $\mathcal{M}_{\textit{surgeon}}$ would rewrite the context rather than the weights, and the weight-level update studied here does not apply.


%% ============================================================
\section{On the Necessity of the Retain Buffer in Machine Unlearning}\label{sec:disc_retain}
%% ============================================================

Effective unlearning requires distinguishing what to forget from what to retain. This section investigates whether the forget data $\mathcal{D}_f$ or the retain buffer $\mathcal{D}_r$, a subset of the full retain data $\mathcal{D}_r^{\textit{full}}$, can be avoided, drawing on evidence from classification, face recognition, and generative tasks.

We find that $\mathcal{D}_r$ is fundamentally unavoidable for preserving task-critical knowledge in non-convex models. Without architectural support such as modular pathways, all parameters are adjusted using both $\mathcal{D}_f$ and $\mathcal{D}_r^{\textit{full}}$ during pretraining, making disentanglement infeasible without access to $\mathcal{D}_r$.

Only noise-based impair-and-repair methods~\citeauthor{tarun2023fast} and source-free unlearning~\citeauthor{ahmed2025towards} attempt to bypass data dependencies. The former still requires $\mathcal{D}_r$ during the repair phase. Source-free methods estimate retain-data Hessians using only $\mathcal{D}_f$ through semi-definite programming, but are limited to convex losses and linear classifiers, making them inapplicable to modern non-convex architectures such as transformers. In practice, they use frozen pre-trained feature extractors and unlearn only linear heads. While error bounds improve with dimensionality, Hessian storage and inversion become computationally prohibitive for large models.

All established methods for non-convex models critically rely on $\mathcal{D}_r$. Weight-importance methods such as EWC~\citeauthor{kirkpatrick2017overcoming} and SALUN~\citeauthor{fan2023salun} compute Fisher information or saliency from $\mathcal{D}_r$ to protect retain-relevant weights. SCRUB~\citeauthor{kurmanji2023towards} and GS-LoRA~\citeauthor{zhao2024continual} explicitly leverage $\mathcal{D}_r$ to safeguard retained knowledge. Continual learning-based unlearning methods universally depend on $\mathcal{D}_r$, where LSF~\citeauthor{shibata2021learning} uses mnemonic codes, UnCLe~\citeauthor{adhikari2025unlearning} employs task embeddings, CLPU-DER++~\citeauthor{liu2022continual} maintains experience replay, and UniCLUN~\citeauthor{chatterjee2024unified} and UG-CLU~\citeauthor{huang2025unified} incorporate $\mathcal{D}_r$ for stability.

Generative tasks present partial exceptions. FLAT~\citeauthor{wang2024llm} solves unlearning without $\mathcal{D}_r$ through f-divergence regularizers, while ESD~\citeauthor{gandikota2023erasing} avoids both $\mathcal{D}_f$ and $\mathcal{D}_r$ by using teacher models to generate synthetic samples. However, FLAT remains limited to batch-level forgetting and ESD requires access to a teacher model.

In summary, while $\mathcal{D}_f$ can theoretically be avoided through noise-based proxies or source-free methods, no viable approach eliminates $\mathcal{D}_r$ dependence for non-convex deep learning without compromising retention performance. This finding contextualizes STAMP's retain buffer requirement not as a limitation specific to our method, but as a fundamental constraint of the machine unlearning problem in its current form.