%============================ack.tex=====================================
%
\begin{Acknowledgements}	

I would like to express my sincere gratitude to everyone who has supported and guided me throughout my M.Tech (R) journey.

First and foremost, I am deeply indebted to my supervisor, \textbf{Dr. Amit Shukla}, and co-supervisor, \textbf{Dr. Praful Hambarde}. Their unwavering support, patience, and wisdom played a pivotal role in shaping this thesis and inspired me to strive for excellence in my research.

I extend my thanks to my \textbf{APC members} for their valuable suggestions and constructive feedback, which greatly improved the quality of this work.

My heartfelt appreciation goes to my labmates, whose camaraderie made this experience truly memorable. I am particularly grateful to my seniors \textbf{Ritali Vatsi Gupta}, \textbf{Ashok Kumar}, \textbf{Surya Prakash}, \textbf{Pranav Singh} and \textbf{Naisarg Pandya} for their mentorship. I also thank my colleagues \textbf{Ashish}, \textbf{Ankit}, \textbf{DarshanKumar}, \textbf{Pranav}, \textbf{Shwetha}, and \textbf{Vikas} for their constant encouragement and cooperation.

I also acknowledge the invaluable technical assistance provided by our lab staff, \textbf{Mr. Abhay}, \textbf{Mr. Shivkamal}, \textbf{Mr. Prashant}, \textbf{Mr. Aniket}, and \textbf{Sahil}, whose dedication ensured smooth research operations.

My time at IIT Mandi would not have been as joyful without my friends \textbf{Isha} and \textbf{Shivam}, thank you for making this journey fulfilling.


%
%
\signature[Indian Institute of Technology Mandi]{\today}
\end{Acknowledgements}
%========================================================================
% For plain page style, use below environment-
%\begin{Plainacknowledgements}
%	contents ............
%\end{Plainacknowledgements}


%%% Local  Variables: 
%%% mode:  latex 
%%% TeX-master: "Thesis.tex"  
%%% End: 











